Classless Java: Tuning Java Interfaces

### 2015-12-08 paper/sections/CaseStudy.tex
manual check

\noindent The main benefit that fluent methods give
us is the convinience of method chaining:

@@

\paragraph{Extending the Query Language} The previous query language can be extended with a new feature
\texttt{orderBy} which orders the result records by a field that users
specify. With \mixin programmers just need to extend the interface \texttt{Database} with new
features, and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=14-16]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT
This way, when a \texttt{ExtendedDatabase} query created, 
all the fluent setters return the correct type, and not the old \texttt{Database} type, which would prevent calling 
\texttt{orderBy}.

>>>>

\noindent The main benefit that fluent methods give
us is the convenience of method chaining:

@@

\paragraph{Extending the Query Language} The previous query language can be extended with a new feature
\texttt{orderBy} which orders the result records by a field that users
specify. With \mixin programmers just need to extend the interface \texttt{Database} with new
features, and the return type of fluent setters in
\texttt{Database} is automatically refined to \texttt{ExtendedDatabase}:
\lstinputlisting[linerange=14-16]{../UseMixinLombok/src/jooq/DatabaseTest.java}% APPLY:linerange=FLUENT_DATABASE_EXT
In this way, when a query created using \texttt{ExtendedDatabase}, 
all the fluent setters return the correct type, and not the old \texttt{Database} type, which would prevent calling 
\texttt{orderBy}.

### 2015-12-08 paper/sections/Formalization.tex
Added a newline for formatting
remove @ObjOf in grammar

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\weakAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of Classless Java}
\label{Grammar}
\end{figure}

@@

%\paragraph{\shadow{}} chooses the most specific version of a method,
%that is the unique version available, or a conflicted version from a
%set of possibilities. In this case all the conflicting defintions are returned.
%We do not model overloading, so it is an error if multiple versions are available with different parameter types. Formally:
%
%\noindent
%$\begin{array}{l}
%\InTextDef{15ex}{
%\shadow()}{\none}\\
%\InTextDef{15ex}{\shadow(\method)}{\method}\\
%\InTextDef{15ex}{\shadow(\overline{\mh\QM;})}{\Aux{mostSpecific}(\overline{\mh\QM;})}\\
%\InTextDef{15ex}{\shadow(\methods)}{\methods}\\
%\InTextWith{\mbox{ the former case not applicable} }
%%\methods\mbox{ not of the form }\overline{\mh\QM;}
%%\mbox{ and }
%%\Aux{mostSpecific}(\methods)\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}\}}\\
%\end{array}$
%
\paragraph{override} models how a method in an interface can
override an implementation in the superinterface, even in case of a
conflict. Note how the special value $\none$ is used, and how (the 4th
case) overriding can solve a conflict.
% The notation in the last case below
%means $\method'$ is not static, and is supertype of $\method$.

\noindent$\begin{array}{l}
\InTextDef{20ex}{\override(\none,\emptyset)}{\none}\\
\InTextDef{20ex}{\override(\method,\emptyset)}{\method}\\
\InTextDef{20ex}{\override(\none,\method)}{\method}\\
\InTextDef{20ex}{\override(\none,\overline{\mh\QM;})}{\Aux{mostSpecific}(\overline{\mh\QM;})}\\
\InTextDef{20ex}{\override(\method,\methods)}{\method}\\
\InTextWith{
\forall \method' \in \methods : 
%\method'\in\{\mh\QM;,\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}, \conflicted\ \mh\QM; \},
\method\subtype\method'
}
\end{array}$
\yanlin{added a colon}
\noindent The definition $\Aux{mostSpecific}$ returns the most
specific method whose type is the subtype of all the
others.  Since method subtyping is a partial ordering,
$\Aux{mostSpecific}$ may not be defined, this in turn forces us to rely on the last clause of $\override$; otherwise
the whole \mBody{} would not be defined for that specific $\m$.
Rule \rn{t-intf} relies on this behaviour.

@@

%The \textsf{shadow} function takes two same methods (with the same name and types of arguments), and return the method which shadows the other during inheritance.
%exmples to motivate our design
%interface A{static String m(){return "A";}}
%interface C extends A{
%	default String dm(){
%		  this.m();//wrong in java
%		  A.m();
%		  C.m();//wrong in java
%		}
%}
%
%
%(1) Static methods are not inherited. Also, if one of $\{body_1,body_2\}$ is null, \textsf{shadow} simply returns the other one. Hence

>>>>

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of Classless Java}
\label{Grammar}
\end{figure}

@@

%\paragraph{\shadow{}} chooses the most specific version of a method,
%that is the unique version available, or a conflicted version from a
%set of possibilities. In this case all the conflicting defintions are returned.
%We do not model overloading, so it is an error if multiple versions are available with different parameter types. Formally:
%
%\noindent
%$\begin{array}{l}
%\InTextDef{15ex}{
%\shadow()}{\none}\\
%\InTextDef{15ex}{\shadow(\method)}{\method}\\
%\InTextDef{15ex}{\shadow(\overline{\mh\QM;})}{\Aux{mostSpecific}(\overline{\mh\QM;})}\\
%\InTextDef{15ex}{\shadow(\methods)}{\methods}\\
%\InTextWith{\mbox{ the former case not applicable} }
%%\methods\mbox{ not of the form }\overline{\mh\QM;}
%%\mbox{ and }
%%\Aux{mostSpecific}(\methods)\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}\}}\\
%\end{array}$
%
\paragraph{override} models how a method in an interface can
override implementations in its superinterfaces, even in the case of
conflicts. Note how the special value $\none$ is used, and how (the 4th
case) overriding can solve a conflict.
% The notation in the last case below
%means $\method'$ is not static, and is supertype of $\method$.

\noindent$\begin{array}{l}
\InTextDef{20ex}{\override(\none,\emptyset)}{\none}\\
\InTextDef{20ex}{\override(\method,\emptyset)}{\method}\\
\InTextDef{20ex}{\override(\none,\method)}{\method}\\
\InTextDef{20ex}{\override(\none,\overline{\mh\QM;})}{\Aux{mostSpecific}(\overline{\mh\QM;})}\\
\InTextDef{20ex}{\override(\method,\methods)}{\method}\\
\InTextWith{
\forall \method' \in \methods :
%\method'\in\{\mh\QM;,\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}, \conflicted\ \mh\QM; \},
\method\subtype\method'
}
\end{array}$

\yanlin{added a colon}
\noindent The definition $\Aux{mostSpecific}$ returns the most
specific method whose type is the subtype of all the
others.  Since method subtyping is a partial ordering,
$\Aux{mostSpecific}$ may not be defined, this in turn forces us to rely on the last clause of $\override$; otherwise
the whole \mBody{} would not be defined for that specific $\m$.
Rule \rn{t-intf} relies on this behaviour.

@@

\paragraph{Example}
To illustrate the mechanism of $\mBody$ in a clearer and more intuitive way, we present an example here.
\begin{lstlisting}
interface A { Object m(); }
interface B extends A { default Object m() {return new Object(){};} }
interface C extends A {}
interface D extends B, C { String m(); }
\end{lstlisting}
We start to compute $\mBody(\QM{m},\QM{D})$ inside out, and step by step:
\begin{itemize}
\item First the full set of supertypes of $\QM{D}$ is obtained recursively and results in $\{\QM{A,B,C}\}$.
\item To compute $\tops(\QM{m},\{\QM{A,B,C}\})$, $\QM{C}$ can be omitted, since $\QM{m}$ is not declared directly in $\QM{C}$, hence $\Aux{originalMethod}(\QM{m},\QM{C})$ is undefined. And $\QM{B}$ is a subtype of $\QM{A}$ (thus $\QM{B.m}$ hides $\QM{A.m}$), so $\tops(\QM{m},\{\QM{A,B,C}\})$ equals $\QM{B.m}$.
\item Now the final step is to compute $\override(\QM{D.m},\QM{B.m})$, and this concludes by the fifth case of $\override$ that $\QM{D.m}$ hides $\QM{B.m}$ successfully ($\QM{String}$ is a subtype of $\QM{Object}$) and becomes the most specific method. So finally we get $\mBody(\QM{m},\QM{D}) = \QM{D.m}$.
\end{itemize}
%The \textsf{shadow} function takes two same methods (with the same name and types of arguments), and return the method which shadows the other during inheritance.
%exmples to motivate our design
%interface A{static String m(){return "A";}}
%interface C extends A{
%	default String dm(){
%		  this.m();//wrong in java
%		  A.m();
%		  C.m();//wrong in java
%		}
%}
%
%
%(1) Static methods are not inherited. Also, if one of $\{body_1,body_2\}$ is null, \textsf{shadow} simply returns the other one. Hence

### 2015-12-08 paper/sections/Introduction.tex
text driving Marco crazy

\item {\em Interfaces do not have state.} This creates a tension between 
 using multiple inheritance and having state. Using setter and
  getter methods is a way out of this tension, but this workaround
  requires tedious boilerplate classes that latter implement those
  methods.

@@

Object interfaces do not require changes to the Java runtime or compiler, 
and they also do not introduce any new syntax (apart from a
small annotation). All three features of object interfaces are
achieved by reinterpreting existing Java syntax, and are translated
into regular Java code without loss of type-safety. Since no new
syntax is introduced, it would be incorrect to call object interfaces
a language extension or syntactic sugar. So we use the term
\emph{language tuning} instead. Language tuning sits in between a
lightweight language extension and a glorified library. Language
tuning can offer many features usually implemented by a real language
extension, but because it does not modify the language syntax
pre-existing tools can work transparently on the tuned language.  To
exploit the full-benefits of language tuning, our prototype
implementation of object interfaces uses Java annotations and AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

>>>>

\item {\em Interfaces do not have state.} This creates a tension between 
 using multiple inheritance and having state. Using setter and
  getter methods is a way out of this tension, but this workaround
  requires tedious boilerplate classes that later implement those
  methods.

@@

Object interfaces do not require changes to the Java runtime or compiler, 
and they also do not introduce any new syntax. All three features of object interfaces are
achieved by reinterpreting existing Java syntax, and are translated
into regular Java code without loss of type-safety. Since no new
syntax is introduced, it would be incorrect to call object interfaces
a language extension or syntactic sugar. So we use the term
\emph{language tuning} instead. Language tuning sits in between a
lightweight language extension and a glorified library. Language
tuning can offer many features usually implemented by a real language
extension, but because it does not modify the language syntax
pre-existing tools can work transparently on the tuned language.  To
exploit the full-benefits of language tuning, our prototype
implementation of object interfaces uses Java annotations and AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

### 2015-12-08 paper/sections/Overview.tex
removed some line breaks

This section illustrates how object interfaces, expressed using the \mixinAnn{} annotation, enables powerful
idioms using multiple inheritance in Java.
To propose a standard example for multiple inheritance, we show \Q@Animal@s with a two-dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their locations need to be three-dimensional \Q@Point3D@s.
Finally, we model \Q@Pegasus@ (one of the best-known creatures in Greek
mythology) as a kind of \Q@Animal@ with the skills
of both \Q@Horse@s and \Q@Bird@s. A simple class diagram illustrating
the basic system is given in the left side of Figure~\ref{fig:pegasus}.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argues against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

@@

\noindent If an implementation is provided for methods,
only code for methods needing implementation is generated. Thus,
programmers can easily tailor the behaviour for their special needs.
%\end{comment}

@@

\paragraph{Unsatisfactory class-based solutions to field type refinement}
In Java if we define an animal class with a field we have a set of
unsatisfactory options in front of us:

\begin{itemize}

\item Define a \Q@Point3D@ field in \Q@Animal@: this is bad since all animals
  would require more than needed.
  %Also it requires the programmer to predict the future, or
  Also it requires adapting the old code to accommodate for new evolutions.

@@

\noindent \Q@Point3D@ includes a
\Q@with@ method, taking a \Q@Point2D@ as an argument.
Other wither methods (such as \Q@withX@) functionally update a field at a time.  This can be
inefficient, and sometimes hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object with all fields that match the fields of updated
parameter type. The idea is that the result should be like \Q@this@, but
modified to be as similar as possible to the parameter.

>>>>

This section illustrates how object interfaces, expressed using the \mixinAnn{} annotation, enables powerful
idioms using multiple inheritance in Java.
To propose a standard example for multiple inheritance, we show \Q@Animal@s with a two-dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their locations need to be three-dimensional \Q@Point3D@s.
Finally, we model \Q@Pegasus@ (one of the best-known creatures in Greek
mythology) as a kind of \Q@Animal@ with the skills
of both \Q@Horse@s and \Q@Bird@s. A simple class diagram illustrating
the basic system is given in the left side of Figure~\ref{fig:pegasus}.\footnote{Some research argues in favour of using subtyping for modelling taxonomies, other research argues against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

@@

\noindent Only code for methods needing implementation is generated. Thus,
programmers can easily customize the behaviour for their special needs.
%\end{comment}

@@

\paragraph{Unsatisfactory class-based solutions to field type refinement}
In Java if we define an animal class with a field we have a set of
unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@: this is bad since all animals
  would require more than needed.
  %Also it requires the programmer to predict the future, or
  Also it requires adapting the old code to accommodate for new evolutions.

@@

\noindent \Q@Point3D@ includes a
\Q@with@ method, taking a \Q@Point2D@ as an argument.
Other wither methods (such as \Q@withX@) functionally update a field at a time.  This can be
inefficient, and sometimes hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object where all the fields that match fields in the parameter
object are updated to the corresponding value in the parameter. The idea is that
the result should be like \Q@this@, but modified to be as similar as possible to the parameter.

### 2015-12-08 paper/sections/Translation.tex

This section shows what the \mixin annotation generates and presents a
formal definition for most of the generated methods. Since
Classless Java does not consider
% setters, so we do not formally define the generation of the setters.
%We do not include
casts or \Q@instanceof@, the \Q@with@ method is not included in the
formal translation. For the same reason \Q@void@ returning setters are
not included, since they are just a minor variation over the more
interesting fluent setters, and they would require special handling
just for the conventional \Q@void@ type.

\subsection{Translation}

For the purposes of the formalization, the translation is divided into
two parts for more convenient discussion on formal properties later. To this aim we introduce the annotation
$\weakAnn$. Its role is only in the translation process, hence is
not part of the Classless Java language.  $\weakAnn$ generates the
constructor method $\QM{of}$, while \mixin automatically refines the
return types and calls $\weakAnn$.

@@

In the following sections  we will write $\QM{with#}\m$ to append $\m$ to
\QM{with}, following the camelCase rule. The first letter of $\m$
must be lower-case and is changed to upper-case upon appending. For
example \QM{with#foo}=\QM{withFoo}.  Special names $\specialName(\m)$
are \QM{with} and all identifiers of the form $\QM{with#}\m$.

\paragraph{The $\otherMethod$ method:}
$\otherMethod(\C_0,\methods)$ is defined as follows:

@@

\paragraph{The $\ofMethod$ method:}\label{subsec:ofmethod}
The function $\ofMethod$ generates the method \QM{of}, as an object factory. To
avoid boring digressions into well-known ways to find unique
names, %for the sake of this formalization
we assume that all methods with no parameters do not start with an underscore,
and we prefix method names with underscores to obtain valid parameter names for
\QM{of}.

@@

\noindent$\begin{array}{ll}
\InTextDef{24ex}{\method\in\fieldsFunc(\C_0)}{
\isField(\method)\ \mand\
%}\\\InTextWith{
\method=\mBody(\_,\C_0)
}
\end{array}$

\noindent For methods inside the interface with the form $\C_i\ \m_i$\QM{();}
  \begin{itemize}
   \item $\m_i$ is the field name, and has type $\C_i$.
   \item $\m_i$\QM{()} is the getter, and just returns the current field value.
   \item if a method \Q@with#@$\m_i$ is required, then it is implemented by calling the \Q@of@ method using
    the current value for all the fields except for $\m_i$. Such new value is
    provided as a parameter. This corresponds to the expressions $\es_i$.
\item \QM_$\m_i$\QM($\C_i\ $\QM{ _val)} is the setter. In our prototype we use name $\m_i$, here we use the underscore to avoid modelling overloading.
%   \item similarly, for the \Q@clone@ method, \Q@of@ is called using the current value for all the fields.
   %\item To complete our generation, we need to generate setters, fluent setters and the with method.
   %\item \marco{should we just formalize setters?}
   \end{itemize}

>>>>

This section shows what the \mixin annotation generates and presents a
formal definition for most of the generated methods. Since the formalized part
of Classless Java does not consider
% setters, so we do not formally define the generation of the setters.
%We do not include
casts or \Q@instanceof@, the \Q@with@ method is not included in the
formal translation. For the same reason \Q@void@ returning setters are
not included, since they are just a minor variation over the more
interesting fluent setters, and they would require special handling
just for the conventional \Q@void@ type.

\subsection{Translation}
For the purposes of the formalization, the translation is divided into
two parts for more convenient discussion on formal properties later. To this aim we introduce the annotation
$\weakAnn$. Its role is only in the translation process, hence is
not part of the Classless Java language.  $\weakAnn$ generates the
constructor method $\QM{of}$, while \mixin automatically refines the
return types and calls $\weakAnn$.

@@

We write $\QM{with#}\m$ to append $\m$ to
\QM{with}, following the camelCase rule. The first letter of $\m$
must be lower-case and is changed to upper-case upon appending. For
example \QM{with#foo}=\QM{withFoo}.  Special names $\specialName(\m)$
are \QM{with} and all identifiers of the form $\QM{with#}\m$.

\paragraph{The $\otherMethod$ function:}
$\otherMethod(\C_0,\methods)$ is defined as follows:

@@

\paragraph{The $\ofMethod$ function:}\label{subsec:ofmethod}
The function $\ofMethod$ generates the method \QM{of}, as an object factory. To
avoid boring digressions into well-known ways to find unique
names, %for the sake of this formalization
we assume that all methods with no parameters do not start with an underscore,
and we prefix method names with underscores to obtain valid parameter names for
\QM{of}.

@@

\noindent$\begin{array}{ll}
\InTextDef{24ex}{\method\in\fieldsFunc(\C_0)}{
\isField(\method)\ \mand\
%}\\\InTextWith{
\method=\mBody(m^\method,\C_0)
}
\end{array}$

\noindent For methods inside the interface with the form $\C_i\ \m_i$\QM{();}
  \begin{itemize}
   \item $\m_i$ is the field name, and has type $\C_i$.
   \item $\m_i$\QM{()} is the getter, and just returns the current field value.
   \item if a method \Q@with#@$\m_i()$ is required, then it is implemented by calling the \Q@of@ method using
    the current value for all the fields except for $\m_i$. Such new value is
    provided as a parameter. This corresponds to the expressions $\es_i$.
\item \QM_$\m_i$\QM($\C_i\ $\QM{ _val)} is the setter. In our prototype we use name $\m_i$, here we use the underscore to avoid modelling overloading.
%   \item similarly, for the \Q@clone@ method, \Q@of@ is called using the current value for all the fields.
   %\item To complete our generation, we need to generate setters, fluent setters and the with method.
   %\item \marco{should we just formalize setters?}
   \end{itemize}
